---
title: "Policy Evaluation during a Pandemic"
subtitle: ""
author: "Brantly Callaway$^1$ and Tong Li$^2$<br><br> $^1$University of Georgia, $^2$Vanderbilt University <br><br><br>"
institute: ""
date: "November 20, 2022<br>Southern Economics Association Conference"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["metropolis_alt.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---

# Introduction


$$\newcommand{\E}{\mathbb{E}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathrm{var}}
\newcommand{\cov}{\mathrm{cov}}
\newcommand{\Var}{\mathrm{var}}
\newcommand{\Cov}{\mathrm{cov}}
\newcommand{\Corr}{\mathrm{corr}}
\newcommand{\corr}{\mathrm{corr}}
\newcommand{\L}{\mathrm{L}}
\renewcommand{\P}{\mathrm{P}}
\newcommand{\independent}{{\perp\!\!\!\perp}}$$


<style type="text/css">

border-top: 80px solid #BA0C2F;

.inverse {
  background-color: #BA0C2F;
}

.alert {
    font-weight:bold; 
    color: #BA0C2F;
}

.alert-blue {
    font-weight: bold;
    color: blue;
}

.remark-slide-content {
    font-size: 23px;
    padding: 1em 4em 1em 4em;
}

.highlight-red {
 background-color:red;
 padding:0.1em 0.2em;
}

.assumption-box {
    background-color: rgba(222,222,222,.5);
    font-size: x-large;
    padding: 10px; 
    border: 10px solid lightgray; 
    margin: 10px;
}

.assumption-title {
    font-size: x-large;
    font-weight: bold;
    display: block;
    margin: 10px;
    text-decoration: underline;
    color: #BA0C2F;
}
</style>


```{r echo=FALSE}
step_by_step_eq <- function(eqlist, before="", after="", title=" ") {
  
  # drop slide pauses in before content
  before_inner <- gsub("--", "", before)
  
  for (i in 2:length(eqlist)) {
    eqlist[i] <- paste0(eqlist[i-1],"\\\\\n",eqlist[i])
  }

  out <- ""
  for (i in 1:length(eqlist)) {
    if (i > 1) out <- paste0(out, "count:false", "\n")
    out <- paste0(out, "# ", title, "\n") # print title
    # print before content
    if (i == 1) out <- paste0(out, before, "\n") else out <- paste0(out, before_inner, "\n") 
    out <- paste0(out, "$$\n\\begin{aligned}\n",eqlist[[i]],"\n\\end{aligned}\n$$\n\n") # print equation
    if (i < length(eqlist)) out <- paste0(out, "---\n\n")
  }
  out <- paste0(out, after, "\n") # print after content
  out <- paste0(out, "---\n")
  cat(out)
}
```

- Tons of recent empirical work evaluating the effects of Covid-related policies

    - Shelter-in-place orders, school closings, mask mandates, contact tracing, testing policies, etc.

    - Recent summary of the literature: Haber et al. (2021)

--

- At first glance, this seems <span class="alert">in the wheelhouse</span> of economists

    - Large amount of relevant data
    
    - Policies enacted at different points in time across different locations
    
--
    
    - $\implies$ Difference-in-differences
    
---

# Introduction

In this paper, we are going to take a <span class="alert">second look</span>

--

1. Hard to rationalize DID at all

    - Not generally compatible with leading pandemic models
    
    - Pre-testing does not work well (especially early in the pandemic)
    
--

2. We suggest using a particular version of <span class="alert">unconfoundedness</span> that involves conditioning on important pre-treatment pandemic related variables

    - We'll largely ignore a number of other potential challenges: e.g., measurement error in cases, multiple policies, spillovers, non-random/limited Covid-19 testing
    
--

3. We also propose an approach to study effects of Covid-related policies on <span class="alert">economic outcomes: </span> e.g., unemployment, travel, etc.

---



# Outline

<br> <br> <br>

1. Stochastic SIRD Models

2. Policy Effects on Covid-19 Cases

3. Application: Shelter-in-Place Orders

---

class: inverse, middle, center
count: false

# Stochastic SIRD Models

---

# Stochastic SIRD Model

SIRD models are the workhorse model for the spread of an epidemic

- Kermack and McKendrick (1927), Allen (2008), and Allen (2017)

--

- In economics: Oka, Wei, and Zhu (2020), Fernandez-Villaverde and Jones (2020), Ellison (2020), Acemoglu, Chernozhukov, Werning, and Whinston (2020), others

--

In a particular location, all individuals are in one of four states:

--

S - Susceptible

--

I - Infected

--

R - Recovered

--

$\delta$ - Dead

---

# Stochastic SIRD Model


Additional Notation:

- $N_l$ - number of individuals in location $l$

- $C_{lt}$ - cumulative cases in location $l$ by time period $t$

- $\mathcal{F}_{lt} = (S_{lt}, I_{lt}, R_{lt}, \delta_{lt})$ - "state" of the pandemic in location $l$ in time period $t$

--

Key idea: Markov transition equations for active cases, recoveries, deaths, number susceptible, and cumulative cases

--

Additional Parameters:

- $\lambda$ - recovery rate

- $\gamma$ - death rate

- $\beta$ - infection rate

---

# Stochastic SIRD Model

<span class="alert">Active Cases:</span>

$$\E[I_{lt} | \mathcal{F}_{lt-1}] = (1-\lambda-\gamma) I_{lt-1} + \beta \frac{I_{lt-1}}{N_l} S_{lt-1}$$

--

<span class="alert">Recoveries:</span>

$$\E[R_{lt} | \mathcal{F}_{lt-1}] = R_{lt-1} + \lambda I_{lt-1}$$

--


<span class="alert">Deaths:<span>

$$\E[\delta_{lt} | \mathcal{F}_{lt-1}] = \delta_{lt-1} + \gamma I_{lt-1}$$

---

# Stochastic SIRD Model 

<span class="alert">Susceptible:</span>

$$\E[S_{lt}| \mathcal{F}_{lt-1}] = S_{lt-1} - \beta \frac{I_{lt-1}}{N_l} S_{lt-1}$$

--

<span class="alert">Cumulative Cases:</span>

$$\E[C_{lt}| \mathcal{F}_{lt-1}] = C_{lt-1} + \beta \frac{I_{lt-1}}{N_l}S_{lt-1}$$

---

# Simulated SIRD Pandemic

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("sim_example.pdf", density=83)
```

---

class: inverse, middle, center
count: false

# Policy Effects on Covid-19 Cases

---

# Setup

<span class="alert">Additional Potential Outcomes Notation:</span>

--

- $D_l$ - indicates whether or not location $l$ participates in the treatment

--

- Treated potential outcomes: $S_{lt}(1)$, $I_{lt}(1)$, $R_{lt}(1)$, $\delta_{lt}(1)$, $C_{lt}(1)$

--

- Untreated potential outcomes: $S_{lt}(0)$, $I_{lt}(0)$, $R_{lt}(0)$, $\delta_{lt}(0)$, $C_{lt}(0)$

--

- $\mathcal{T}$ total time periods and treatment occurs in time period $t^*$.

    - For simplicity, suppose that treatment timing is the same across all locations
    
--

- Observed outcomes (for example):

    - In pre-treatment periods, when $t < t^*$: $C_{lt} = C_{lt}(0)$ for all locations 
    
    - In post-treatment periods, when $t \geq t^*$: $C_{lt} = D_l C_{lt}(1) + (1-D_l) C_{lt}(0)$

--

- Long difference notation, for $t_2 > t_1$: $\Delta^{(t_1, t_2)} C_t := C_{t_2} - C_{t_1}$

---

# Target Parameter

- Target parameter, for $t \geq t^*$:

$$ATT^C_t = \E[C_t(1) - C_t(0) | D=1]$$

---

# Review of Difference-in-Differences

The main assumption underlying a DID approach is the <span class="alert">parallel trends assumption:</span>

--

## Parallel Trends Assumption

For all $t=2,\ldots,\mathcal{T}$,

$$\E[\Delta C_t(0) | D=1] = \E[\Delta C_t(0) | D=0]$$

--

DID estimand given by:

$$DID^C_t = \E[\Delta^{(t^*-1,t)}C_t | D=1] - \E[\Delta^{(t^*-1,t)} C_t |D=0]$$

--

Under parallel trends, it essentially immediately follows that

$$ATT^C_t = DID^C_t$$

---

# Difference-in-Differences

DID is very closely related to the following model for untreated potential outcomes:

$$C_{lt}(0) = \theta_t + \eta_l + v_{lt}$$

--

(see, for example, Gardner (2021), Borusyak, Jaravel, and Spiess (2021), Ghanem, Sant'Anna, and Wuthrich (2022)) and where

--

- $\theta_t$ - is a time fixed effect

- $\eta_l$ - is a location (individual) fixed effect (can be distributed differently for the treated group relative to )

- $v_{lt}$ - idiosyncratic unobservables

--

Taking difference $C_{lt}(0)$ over time gets rid of $\eta_l$ and immediately implies parallel trends holds



---


# Difference-in-Differences

DID is very closely related to the following model for untreated potential outcomes:

$$C_{lt}(0) = \theta_t + \eta_l + v_{lt}$$

--

<span class="alert">This is a much different model from the SIRD model we discussed earlier</span>

--

$$C_{lt}(0) = C_{lt-1}(0) + \beta \frac{I_{lt-1}(0)}{N_l}S_{lt-1}(0) + v_{lt}$$

--

Side-comment: You can re-arrange this to get 

$$\Delta C_{lt}(0) = \beta \frac{I_{lt-1}(0)}{N_l}S_{lt-1}(0) + v_{lt}$$

but $S_{lt-1}(0) = N_l - C_{lt-1}(0) \implies$ need to condition on pre-treatment outcome $\stackrel{\textrm{in my opinion}}{\implies}$ no longer DID because it reduces to unconfoundedness (see Abadie (2005))

---

# Main Results

Under a stochastic SIRD model for untreated potential outcomes:

--

1. Parallel trends does not generally hold

--

2. The bias from incorrectly imposing parallel trends is:

--

    - Potentially complicated
  
    - Depends on differences in the distribution of the "state" of the pandemic in period $t^*-1$ between the treated and untreated group

    - Cannot "sign" the bias in general, but suppose that (i) early policy adopters have more pre-policy Covid-19 cases, and (ii) looking at short-term effects of policy $\implies$ may spuriously estimate *positive* effects of policies
    
--

3. A version of unconfoundedness holds (mainly: conditional on the pre-treatment "state" of the pandemic)
    

---

# Complications

There are a number of ways you could make things more complicated

--

Some are compatible with our framework:

--

- SIRD parameters that vary over time

--

- SIRD parameters that depend on a location's observed characteristics (e.g., region)

--

- Anticipation effects 

--

Some are not (immediately) compatible:

--

- Measurement error in cases

--

- Spillovers

--

- TWFE versions of SIRD parameters, such as: $\beta_{lt} = \beta_l + \beta_t$

---

class: inverse, center, middle
count: false

# Application: Shelter-in-Place Orders

---

# Application Setup 

We study Shelter-in-Place Orders (SIPOs) early in the pandemic, and their efffect on

- Covid-19 cases (from CDC Covid Data Tracker) <span class="alert"><-- This one only today</span>

- Travel (from Google Mobility reports) 

--

These have been widely studied in a number of papers

--

- Consider early part of the pandemic: March 10, 2020 to May 1, 2020

--

- State level data <span class="alert"><-- This one only today</span>

- County-level data

--

Main emphasis for today: (non-) robustness of DID methods in this context

---


# Various Versions of DID

To start with, we'll show results from various ways to implement DID

- Outcomes in levels, parallel trends

- Outcomes in levels, linear trends

- Outcomes in logs, parallel trends

- Outcomes in logs, linear trends

--

We'll also use TWFE regression versions of these as well as "treatment effect heterogeneity robust" approaches from Callaway and Sant'Anna (2021) and Gardner (2021).

---


# TWFE, Levels, Parallel Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("level_twfe_es.pdf", density=100)
```


---

# TWFE, Levels, Linear Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("level_twfe_es_lt.pdf", density=100)
```

---

# TWFE, Logs, Parallel Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("log_twfe_es.pdf", density=100)
```

---

# TWFE, Logs, Linear Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("log_twfe_es_lt.pdf", density=100)
```

---

# CS, Levels, Parallel Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("did_level.pdf", density=100)
```


---

# Gardner, Levels, Linear Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("level_did2s_es_lt.pdf", density=100)
```

---

# CS, Logs, Parallel Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("did_log.pdf", density=100)
```

---

# Gardner, Logs, Linear Trends

```{r echo=FALSE, fig.align='center'}
magick::image_read_pdf("log_did2s_es_lt.pdf", density=100)
```

---

# Discussion

(1) DID results are all over the place...not robust to functional form (e.g., Roth and Sant'Anna (2021))

--

* <span class="alert">none</span> of the specifications here compatible with SIRD model (see also Alcott et al. (2020))

--

(2) Positive effects of SIPOs on Covid-19 cases not reasonable 

--

(3) Its infeasible to decide between approaches based on performance is pre-treatment periods

* None of them are rejected in pre-treatment periods.

---

# Additional Results

<span class="alert">Under unconfoundedness:</span> We do not find strong evidence that SIPOs decreased Covid-19 cases

--

- However 1: What's the counterfactual? - Even among untreated states, there are massive decreases in travel during the time period that we consider

- However 2: Confidence bands are actually pretty wide too (compatible with approximately 30% reduction in cases)

--

<span class="alert">Using county-level data</span> (which has its own advantages and disadvantages), we provide further evidence that:

- DID results are all over the place

- DID can result in relatively large estimates of effects of "placebo policies" in states that never implemented a SIPO

- Unconfoundedness performs better along these dimensions, and tends to result in small or no effect of SIPOs on Covid-19 cases.

---


# Conclusion

We studied policy evaluation strategies for understanding Covid-19 related policies

- Because pandemics are highly nonlinear, the choice of identification strategy can be very important

--

- In particular, DID seems unlikely to perform well here

--

- Other identification strategies like unconfoundedness are at least compatible leading pandemic models (and are likely better choices)

--

- That said, there are still a number of complications and, in my view, evaluating Covid-19 related policies is quite challenging

--

<span class="alert">Five Minute Summary:</span> <a href="https://bcallaway11.github.io/posts/five-minute-pandemic-policy">https://bcallaway11.github.io/posts/five-minute-pandemic-policy</a>

<span class="alert">Code:</span> <a href="https://github.com/bcallaway11/ppe">https://github.com/bcallaway11/ppe</a>